### 2025-04-27 paper/paper.tex
Update paper.tex
basic copyedit

After beginning implementation we decided that CityFlow was not the appropriate tool for this problem, despite our research. While a powerful tool, it struggled to handle the data format provided from OSM data, resulting in slow performance and broken intersections. We then decided moving to the SUMO simulator was a more worthwhile decision, as its efficency has improved over time and documentation is more appropriate for our specific research area.

Once SUMO was installed, we downloaded public data for the city of Starkville Mississippi's road structure using the tool open street map, and used SUMO's netedit to process the data into a useable format. Some changes were made, such as fixing traffic signal placement using junctions, and fixing non-signal based intersection interactions.

@@

In order to properly evaluate the performance of threshold and neural network models, a method of evaluation was needed beyond qualitative methods, such as observation or reivew. As a result we implemented various benchmarks to quantitatively decide the quality of a methods approach.

@@

The next is the average trip time. Becuase we will benchmark similar environments, we can get the average trip times of all combined vehicles in a simulation, and use it as a quantitative performance metric. Quicker trips indicates a successful traffic signal control implementation, as people can get to their end destination quicker, the entire point of driving. It also serves as a benchmark for other metrics, likely indicating shorter queue lengths, less stops, and less time waiting at stops on average.

Our final benchmark involves calculating queue lengths, or the average length of cars in line at an intersection. We can calulate both the max and average, to ensure the model is acting appropriately and not building super long lines to get the average down. A long line indicates that the model has a lot of buildup, and might fill later intersections with traffic, resulting in frustrated drivers and numerous stops if not carefully considered; therefore, it is critical that we minimize both the average queue length and the max queue length with our neural network.

@@

After designing an agent we decided to model the simulation environment in python to actually obtain states and actions for each agent. To do this, we used traci to obtain the states of vehicles on the lanes of traffic heading towards a specific traffic light, and assigned them to the traffic light. We then grabbed the possible state of actions, again assigning it to said traffic signal. This is neccessary for computing optimal actions later. We gather these every 5 seconds in the simulation, which greatly increases performance, and prevents rapidly switching signals during learning.

Our intitial plan was to predict the optimal states of every intersection with a traffic light in Starkville, MS, however, upon implementation we realized our available hardware and timeframe would not allow us to do this. Instead, we opted to only use what we decided were points of interest, or intersections where traffic is particularly backed up. Most of these points were decided by considering the importance of the road, the time spent at intersections, and personal experience. This way we don't waste computational resources at intersections that see little traffic, or do fine directing traffic on their own. Doing so significantly decreases the computational requirements to run the model, and also allows us to focus only on important points.

@@

\subsection{qualitative analysis}

When observing the baseline threshold based traffic signals, it did an acceptable job managing light-medium traffic (2000 vehicles per hour), but saw relative redundance with lights changing while no cars were in the opposite direction, causing congestion on major roads. It felt familiar, with minor traffic on less congested roads, such as backroads connected to highways and highways without traffic stops, but was certainly not as effective as possible.

The MARL Deep Q-learning model saw noteable improvements in terms of responsiveness. When there was no signs of incoming traffic or lines in stoped lanes, it skipped the redunancy of changing the state of traffic lights resulting in a generally smoother flow of traffic. When it did change, it seemed intelligent, considering how long cars have been waiting along with how many cars were waiting. While our model did not allow collisions to be considered, the more continuous flow of traffic saw less emergency stops, and generally safer roads. Analysis also revealed that vehicles would often arrive at their destination sooner than their threshold counterparts, with less time spent on the road generally resulting in safer roads with less congestion.

@@

\subsection{quantitative analysis}

While simple observations provide great insight to the performance of our model, numbers to support our observations are potentially more important. As mentioned earlier, we developed several benchmarks based on the results of a simulation. Each benchmark provides a real insight that indicates the perforamnce of a traffic signal control method without the need to make observations.

@@

\subsection{Analysis of overall results}

>>>>

After beginning implementation we decided that CityFlow was not the appropriate tool for this problem, despite our research. While a powerful tool, it struggled to handle the data format provided from OSM data, resulting in slow performance and broken intersections. We then decided moving to the SUMO simulator was a more worthwhile decision, as its efficiency has improved over time and documentation is more appropriate for our specific research area.

Once SUMO was installed, we downloaded public data for the city of Starkville Mississippi's road structure using the tool open street map, and used SUMO's netedit to process the data into a usable format. Some changes were made, such as fixing traffic signal placement using junctions, and fixing non-signal based intersection interactions.

@@

In order to properly evaluate the performance of threshold and neural network models, a method of evaluation was needed beyond qualitative methods, such as observation or review. As a result we implemented various benchmarks to quantitatively decide the quality of a methods approach.

@@

The next is the average trip time. Because we will benchmark similar environments, we can get the average trip times of all combined vehicles in a simulation, and use it as a quantitative performance metric. Quicker trips indicates a successful traffic signal control implementation, as people can get to their end destination quicker, the entire point of driving. It also serves as a benchmark for other metrics, likely indicating shorter queue lengths, less stops, and less time waiting at stops on average.

Our final benchmark involves calculating queue lengths, or the average length of cars in line at an intersection. We can calculate both the max and average, to ensure the model is acting appropriately and not building super long lines to get the average down. A long line indicates that the model has a lot of buildup, and might fill later intersections with traffic, resulting in frustrated drivers and numerous stops if not carefully considered; therefore, it is critical that we minimize both the average queue length and the max queue length with our neural network.

@@

After designing an agent we decided to model the simulation environment in python to actually obtain states and actions for each agent. To do this, we used traci to obtain the states of vehicles on the lanes of traffic heading towards a specific traffic light, and assigned them to the traffic light. We then grabbed the possible state of actions, again assigning it to said traffic signal. This is necessary for computing optimal actions later. We gather these every 5 seconds in the simulation, which greatly increases performance, and prevents rapidly switching signals during learning.

Our initial plan was to predict the optimal states of every intersection with a traffic light in Starkville, MS, however, upon implementation we realized our available hardware and timeframe would not allow us to do this. Instead, we opted to only use what we decided were points of interest, or intersections where traffic is particularly backed up. Most of these points were decided by considering the importance of the road, the time spent at intersections, and personal experience. This way we don't waste computational resources at intersections that see little traffic, or do fine directing traffic on their own. Doing so significantly decreases the computational requirements to run the model, and also allows us to focus only on important points.

@@

\subsection{Qualitative Analysis}

When observing the baseline threshold-based traffic signals, it did an acceptable job managing light-medium traffic (2000 vehicles per hour), but saw relative redundancy with lights changing while no cars were in the opposite direction, causing congestion on major roads. It felt familiar, with minor traffic on less congested roads, such as backroads connected to highways and highways without traffic stops, but was certainly not as effective as possible.

The MARL Deep Q-learning model saw notable improvements in terms of responsiveness. When there was no signs of incoming traffic or lines in stopped lanes, it skipped the redundancy of changing the state of traffic lights resulting in a generally smoother flow of traffic. When it did change, it seemed intelligent, considering how long cars have been waiting along with how many cars were waiting. While our model did not allow collisions to be considered, the more continuous flow of traffic saw less emergency stops, and generally safer roads. Analysis also revealed that vehicles would often arrive at their destination sooner than their threshold counterparts, with less time spent on the road generally resulting in safer roads with less congestion.

@@

\subsection{Quantitative Analysis}

While simple observations provide great insight to the performance of our model, numbers to support our observations are potentially more important. As mentioned earlier, we developed several benchmarks based on the results of a simulation. Each benchmark provides a real insight that indicates the performance of a traffic signal control method without the need to make observations.

@@

\subsection{Analysis of Overall Results}
